Disentangling linkages between satellite-derived indicators of forest structure and productivity for ecosystem monitoring.

### 2024-01-24 scripts/manuscript/manuscript.qmd
third submission

The Essential Biodiversity Variables (EBV) framework has been proposed
as a monitoring system of standardized, comparable variables that
represents a minimum set of biological information to monitor
biodiversity change at large spatial extents. Six classes of EBVs
(genetic composition, species populations, species traits, community
composition, ecosystem structure and ecosystem function) are defined, a
number of which are ideally suited to observation and monitoring by
remote sensing systems. We used moderate-resolution remotely-sensed
indicators representing two ecosystem level EBV classes (ecosystem
structure and function) to assess their complementarity and redundancy
across a range of ecosystems encompassing significant environmental
gradients. Redundancy analyses found that remote sensing indicators of
forest structure were not strongly related to indicators of ecosystem
productivity (represented by the Dynamic Habitat Indices, DHIs), with
the structural information only explaining 16% of the variation in the
DHIs. Complex metrics of forest structure, such as above ground biomass,
did not contribute additional information over simpler height-based
attributes that can be directly estimated with light detection and
ranging (LIDAR) observations. With respect to ecosystem conditions, we
found that forest types and ecosystems dominated by coniferous trees had
less redundancy between the remote sensing indicators when compared to
broadleaf or mixed forest types. Likewise, higher productivity
environments exhibited the least redundancy between indicators, in
contrast to more environmentally stressed regions. We suggest that
biodiversity researchers continue to exploit multiple dimensions of
remote sensing data given the complementary information they provide on
structure and function focused EBVs, which makes them jointly suitable
for monitoring forest ecosystems.

@@

Ecosystem functions are measures of ecosystem performance that are the
consequence of one or multiple ecosystem processes [@pettorelli2018;
@ecosyste2005]. Within the EBV framework, vegetation phenology --- the
timing and duration of leaf-on conditions --- and ecologically available
energy --- the amount of accessible and useable energy available for
growth in an ecosystem --- are useful predictors of species richness and
abundances at various scales [@chase2002; @radeloff2019; @coops2019;
@razenkova2023]. Ecologically available energy is the amount of energy
produced in an ecosystem, which is sensitive to limitations such as
recent disturbances and the water and nutrient supplies, and is
measurable via productivity [@bonn2004; @nguyen2022]. This differs from
ambient energy in that it goes beyond the climatic conditions of an
area, and concerns the amount of resources available for consumption in
an ecosystem [@hawkins2003]. Clear links have been developed between
remote sensing-derived spectral reflectance, particularly in the
near-infrared and the red parts of the electromagnetic spectrum
[@rouse1974], which relate to pigment concentrations and leaf structure,
respectively, as well as productivity [@yoder1994]. Consequently, the
normalized difference vegetation index (NDVI) and other vegetation
indices have been correlated with productivity measures [@huang2021;
@sims2006]. The relationship however, is typically non-linear,
principally due to saturation in the spectral response at moderate to
high leaf area index values [@myneni1994]. As a result, the strength of
the relationship between spectral indices and productivity depends on
the ecosystem type, land cover type, amount of canopy cover, and
productivity in the region [@shi2017]. Likewise, relationships between
productivity and biodiversity also contain uncertainty, recognizing that
biodiversity is governed by a variety of factors of which productivity
is but one [@skidmore2021]. Despite these uncertainties, remotely sensed
vegetation indices have enabled the assessment of biodiversity patterns
at single time points [@bonn2004; @nagendra2010; @phillips2008], or
through time when using time series data [@berry2007; @radeloff2019;
@coops2019]

@@

bec_tab %>%
  gt() %>%
  cols_label(
    group = "BEC Group",
    zone = "BEC Zone",
    zone_nm = "Full Name",
    ppt = "Average Monthly Precipitation (mm/yr)",
    tmax = "Average Monthly Max Temperature (°C)",
    tmin = "Average Monthly Min Temperature (°C)",
    n = "n"
  ) %>%
  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, and their monthly average climate values for precipitation, maximum temperature, and minimum temperature. Climate data from Wang et al. 2016 averaged for 1991-2020.")
```

@@

### National Terrestrial Ecosystem Monitoring System data products

@@

The forest structure, forest disturbance, and land cover layers are
based on a 30-m best-available-pixel (BAP) composite generated by
Hermosilla et al. @hermosilla2016 using the Composite2Change approach.
The composites were developed for each year from 1984-2019 by selecting
the best available imagery from the available growing season Landsat
observations, removing pixels with clouds, cloud shadows, and haze. The
methodology uses the scoring method from White et al. @white2014 to
select the best pixel for a given year from Landsat-5 Thematic Mapper,
Landsat-7 Enhanced Thematic Mapper Plus, and Landsat-8 Operational Land
Imager imagery. A spectral trend analysis was conducted over these
initial BAP composites using the Normalized Burn Ratio (a vegetation
index) on each pixel, to remove unscreened noise, detect changes, and
temporally interpolate data gaps. This resulted in a gap-free
surface-reflectance composite across Canada [@hermosilla2015]. We use
data from 2015 for all National Terrestrial Ecosystem Monitoring System
products, except forest disturbances, which comprise the period
1985-2020.

@@

#### Forest Structure

The forest structural attributes (canopy height, canopy cover,
structural complexity \[coefficient of variation of height returns\],
basal area, aboveground biomass, and gross stem volume) were derived
from the 30-m spatial resolution dataset generated by Matasci et al.
[@matasci2018; @matasci2018b]. In brief, the method extends structural
attributes derived from a set of lidar acquisitions and field plots
across Canada. These data were generated for the treed pixels of Canada
using the BAP Landsat surface reflectance composite and auxiliary data
(i.e., topography, geography) using a k-Nearest Neighbour imputation
[@matasci2018; @matasci2018b]. Accuracy metrics for imputed forest
structural attributes range from an RMSE of 29.7% (structural
complexity) to 82.3% (gross stem volume) and *R^2^* ranging from 0.125
(structural complexity) to 0.712 (gross stem volume) [@matasci2018;
@matasci2018b].

The DHIs were calculated following Razenkova et al.
@razenkovaMediumresolutionDynamicHabitatInPress for the terrestrial area
of BC. In brief, we used Google Earth Engine [@gorelick2017] to obtain
valid Landsat pixels for the study area across a ten-year time span,
centered on 2015 (2011-2020), filtering out pixels containing shadows,
clouds, and cloud shadows using the QA band derived from the fmask
algorithm [@zhu2012], and then calculated the NDVI index. These
estimates were composited into a synthetic year of monthly data by
calculating the monthly median NDVI value. Finally, images of the sum,
minimum, and coefficient of variation across this synthetic year of NDVI
values were then calculated to represent the cumulative, minimum, and
variation DHIs, respectively.

A summary of the datasets used in this study is presented in Table 2.

@@

To obtain representative sample units of the EBV metrics across the
study area, we implemented stratified random sampling across the union
of BEC zones (Table 1) and forest types (i.e., coniferous, broadleaf,
mixed wood, and wetland treed; see @sec-study-area) of BC. Sample units
were restricted to forested pixels that were surrounded by the same
forest type to reduce the chance of sampling mixed pixels. We also
required that sample units were selected from a homogeneous area, to
reduce uncertainty associated with the input datasets [@shang2020];
homogeneous areas were defined as 3x3-pixel windows with a coefficient
of variation for canopy cover and canopy height below 0.5. A minimum
sampling distance of 1-km was implemented to reduce the effects of
spatial autocorrelation. Sample units that had been disturbed in the
last 35 years were discarded by using the disturbance mask generated for
the forested ecosystems of Canada [@hermosilla2016]. A maximum of 500
sample units were selected from each stratum; in the strata where this
was not achieved, all sampling units meeting the above restrictions were
selected. Sampling was conducted in R version 4.2.2 [@R-base] using the
**sgsR** package [@R-sgsR]. Neighbourhood analyses for the land cover
classes and coefficient of variations of canopy height and cover were
calculated in Python version 3.9.

@@

Variation partitioning is displayed using a Venn diagram in which the
percentage of variance explained by each dataset is in a circle, and the
overlap between circles represents the overlap in variance explained. We
add paths to the Venn diagram to indicate how the structural attributes
contribute to the variation in the DHIs. Because the variation
partitioning results summarize the full multivariate solution, rather
than individual RDA axes, we present variation partitioning results
alongside the path diagrams illustrating the details of the RDA results.
RDA and variation partitioning calculations were done in R [@R-base]
version 4.2.2 using the **vegan** package [@R-vegan]. The code
associated with the processing and analysis is available at
https://github.com/emuise/code-structProdSem.

@@

Overall in BC, the intra-annual patterns of productivity as represented
by the DHI variables were weakly related to forest structure
(@fig-rda-var). Less than 16% of the overall variation in the DHIs was
explained by the forest structural attributes. Due to strong
correlations between the axes of the DHIs, the majority of the variation
was held within the first RDA axis, which had strong loadings for all
DHIs, with positive loadings for the cumulative and minimum DHIs, and
negative loadings for the variation DHI. As such, we identified the RDA1
axis as representing overall productivity. The overall productivity axis
had similar loadings (between 0.24-0.27) for all modelled forest
structural attributes and canopy cover. Canopy height had a smaller
loading (0.1), while structural complexity was negatively correlated
with the overall productivity axis. RDA2 explained much less of the
variation in the DHIs. RDA2 had a large positive loading on the minimum
DHI (0.55), with negative loadings on the cumulative (-0.19) and
variation (-0.39) DHIs. Due to the minimum and variation having the most
influence on RDA2, we identified it as a seasonality axis. Canopy cover
and structural complexity had the strongest (albeit negative; -0.17 and
-0.08, respectively) loadings on the seasonality axis (@fig-rda-var A).
Variation partitioning highlighted that the majority of the explained
variation in the DHIs was due to the primary forest structural
attributes (canopy cover, canopy height, and structural complexity; 9%
variation explained). The overlap between primary and modelled
attributes was 3.5%, and the modelled attributes explained 3.2% of the
variation on their own (@fig-rda-var B).

@@

By forest type, results indicated similar or increased amounts of
variance explained when compared to the overall dataset
(@fig-fracts-bar). By BEC zone, results generally indicated smaller
amounts of variation explained, which may be due to the relatively short
environmental gradients within the zones. The overlap between the
primary and modelled datasets was often responsible for the majority of
the variation explained in the DHIs, except when stratifying by forest
types, where this was only true for the coniferous forests. For
broadleaf, mixed wood, and wetland-treed forest types most of the
explained variation in the DHIs was contributed by the primary forest
structural attributes. A similar pattern was found by BEC zones, where
the variation in the DHIs was principally explained by the overlap
between primary and modelled forest structural attributes, except for
CWH, where no variation was explained by the overlap. The Coastal BEC
group had the lowest amount of variation explained, with Coastal
Douglas-fir having \<1% of variation explained by the structural data.
Overall, the DHIs were decoupled from the forest structural attributes,
as shown by the overall variation explained being under 30% regardless
of data stratification.

@@

Stratification by forest type and ecosystem indicated varying loadings
between the RDA axes and predictor/response variables (@fig-radar). All
DHI variables had high loadings on RDA1 (similar to the overall
dataset), with positive cumulative and minimum DHI loadings, and
negative variation DHI loadings. Colder BEC groups (Montane, North, and
Alpine) had smaller minimum DHI loadings, which were half as large as
the other DHI loadings. As such, we identify the first axis as being
consistently associated with overall productivity. Canopy cover
generally had the highest significant predictor loading on RDA1, except
for in non-coniferous forest types (broadleaf, mixed wood and
wetland-treed). These forest types typically had large negative loadings
between RDA1 and structural complexity. The mixed wood forest and
Coastal BEC group had no significant loadings from modelled forest
structural attributes for RDA1. Examining the predictor loadings for
RDA1 spatially across the province shows a geographic pattern in the
loading strengths of primary forest structural attributes (@fig-fcc A).
The interior (Northern and Southern BEC groups) of the province
generally showed high canopy cover loadings, while the Boreal zones in
the northwest had equal loadings in canopy cover and canopy height.
Coastal zones were driven similarly by canopy height and structural
complexity, with very low or non-significant loadings from canopy cover
(@fig-fcc A; @fig-radar). Modelled forest structural attributes for RDA1
generally show similar loadings across the province (@fig-fcc B).

@@

Only seven of sixteen BEC zones had a second RDA axis, with none being
found in Coastal or Alpine BEC zones (@fig-radar; @fig-fcc). All of the
Montane BEC zones had a secondary axis, while some in the Northern and
Southern interior had a secondary axis. RDA2 often showed the largest
absolute loadings in the minimum DHI, which was frequently, but not
always, negative (@fig-radar; see supplementary material). The Montane
group, which had the largest absolute minimum DHI loading, also had the
largest absolute variation DHI loading. The secondary axis had smaller
loadings in the variation and cumulative DHIs, with only the Southern
BEC group having a lower variation DHI loading than cumulative DHI. RDA2
has lower absolute predictor loadings (maximum of \~0.25) when compared
to RDA1 (maximum of \~0.5), with infrequently significant loadings for
canopy cover and structural complexity (@fig-radar). Among the primary
forest structural attributes, RDA2 had the highest loadings with canopy
height in the Southern group, while the Northern group had slightly
larger structural complexity loadings (@fig-radar; @fig-fcc C). Similar
to RDA1, the modelled forest structural attribute loadings were
generally similar to one another (@fig-radar; @fig-fcc D).

@@

Relationships between forest ecosystem structure and function are
increasingly being examined due to advances in data collection methods
[@ali2019; @radeloff2019; @ali2016]. Prior studies have shown both
theoretical [@knyazikhin1998] and empirical [@atkins2018] relationships
between forest ecosystem structure and function, however, the direction
and significance of these relationships have not been consistent
[@ali2019]. We examined these relationships across the province of BC,
Canada, which has significant environmental gradients
[@valentine1978soil; @pojar1987] by using wall-to-wall forest structural
attributes generated by imputing the lidar-derived attributes with
Landsat best-available-pixel composites [@matasci2018; @matasci2018b]
and Landsat-derived summaries of intra-annual vegetation productivity
[@radeloff2019; @razenkovaMediumresolutionDynamicHabitatInPress]. We
found a weak relationship between these remote sensing indicators of
forest ecosystem structure and function overall across BC (15.7% of
variation explained in the DHIs by forest structural attributes), as
well as within individual BEC zones and forest types found in BC
(@fig-fracts-bar).

Our findings indicate that remote sensing-derived forest ecosystem
structure and forest ecosystem function EBVs are well suited to be used
in a complementary fashion for ecosystem integrity monitoring, despite
both EBVs being derived from surface reflectance from the Landsat series
of satellites [@hansen2021]. While we assessed the complementarity of
ecosystem structure and function, a single pair of EBVs measurable at
the ecosystem scale, other EBV class combinations at finer scales should
also be assessed for complementarity. This is especially relevant in
cases such as ours, where remote sensing datasets such as Landsat can
generate a large number of EBVs in different classes [@skidmore2021]. In
addition, Skidmore et al. [@skidmore2021] highlighted that many
satellite-derived biodiversity products can fall into multiple EBV
classes simultaneously -- e.g., leaf area index being in the ecosystem
structure, ecosystem function, and species trait EBV classes --
depending on the spatial scale of analysis and goal of the monitoring
program. As such, it is relevant to consider which classes each
biodiversity product should belong to, for given spatial and temporal
extents. Other EBVs, such as those found in the community composition
and species population classes, may also be inherently linked due to
being based on the same data sources, even if not observed using remote
sensing platforms.

Further, we assessed the information-richness of primary forest
structural attributes when compared to modelled forest structural
attributes to support indicator selection within a single EBV class. We
found that standard lidar-derived measurements, such as canopy height,
canopy cover, and structural complexity (as proposed by Valbuena et al.
[@valbuena2020]) contribute to the majority of the explained variation
in the DHIs, either independently or in their overlap with the modelled
forest structural attributes (@fig-fracts-bar). This could be expected
as modelled forest structural attributes are generally directly based on
the primary forest structural attributes [@coops2021], however, the
creation of modelled forest structural attributes involves modelling
expertise and processing time overhead [@matasci2018; @matasci2018b;
@duncanson2022]. One advantage of these interpreted products is
potential uptake by a broader set of users, and their ability to be more
closely linked to management actions and monitoring results, as is the
case with forest biomass for climate change mitigation [@duncanson2022].
While either set of forest structural attributes may be most relevant
for a given monitoring task, we note that the modelled structural
attributes do not strongly overlap with structural complexity, and as
such, we generally recommend to use the primary forest structural
attributes.

@@

Forest types with deciduous trees (namely broadleaf and mixed wood) had
large, negative structural complexity loadings when compared to the
other forest structural attributes (@fig-radar), with mixed wood forests
having no modelled structural attributes associated with the DHIs. This
structural complexity-DHIs linkage may be due to understory cover
providing additional photosynthesis in these forest types
[@nilsson2005]. In coniferous forests, on the other hand, the DHIs show
a weaker relationship with structural complexity than all other forest
structural attributes. Notably, the Alpine, Montane, and Northern BEC
groups show decoupling of the three DHIs (@fig-radar), with lower
loadings between the Minimum DHI and RDA1. This decoupling in the DHIs
is potentially due to the near certainty of a pixel containing snow
cover over the course of a year leading to consistently low minimum DHI
values across the ecosystem, and thus to weaker minimum DHI loadings
[@razenkova2020; @beck2006]. These ecosystems with DHI decoupling also
showed relatively strong relationships to canopy cover and weak
relationships with structural complexity. This weak relationship to
structural complexity may be due to the dominance of coniferous forests
in these ecosystems, which have been shown to have lower forest
structural complexity across the globe [@ehbrecht2021], with weaker
relationships between canopy height and structural complexity being
found [@atkins2022].

>>>>

The Essential Biodiversity Variables (EBV) framework has been proposed
as a monitoring system of standardized, comparable variables that
represents a minimum set of biological information to monitor
biodiversity change at large spatial extents. Six classes of EBVs
(genetic composition, species populations, species traits, community
composition, ecosystem structure and ecosystem function) are defined, a
number of which are ideally suited to observation and monitoring by
remote sensing systems. We used moderate-resolution remotely-sensed
indicators representing two ecosystem level EBV classes (ecosystem
structure and function) to assess their complementarity and redundancy
across a range of ecosystems encompassing significant environmental
gradients. Redundancy analyses found that remote sensing indicators of
forest structure were not strongly related to indicators of ecosystem
productivity (represented by the Dynamic Habitat Indices, DHIs), with
the structural information only explaining 15.7% of the variation in the
DHIs. Complex metrics of forest structure, such as aboveground biomass,
did not contribute additional information over simpler height-based
attributes that can be directly estimated with light detection and
ranging (LIDAR) observations. With respect to ecosystem conditions, we
found that forest types and ecosystems dominated by coniferous trees had
less redundancy between the remote sensing indicators when compared to
broadleaf or mixed forest types. Likewise, higher productivity
environments exhibited the least redundancy between indicators, in
contrast to more environmentally stressed regions. We suggest that
biodiversity researchers continue to exploit multiple dimensions of
remote sensing data given the complementary information they provide on
structure and function focused EBVs, which makes them jointly suitable
for monitoring forest ecosystems.

@@

Ecosystem functions are measures of ecosystem performance that are the
consequence of one or multiple ecosystem processes [@pettorelli2018;
@ecosyste2005]. Within the EBV framework, vegetation phenology --- the
timing and duration of leaf-on conditions --- and ecologically available
energy --- the amount of accessible and useable energy available for
growth in an ecosystem --- are useful predictors of species richness and
abundances at various scales [@chase2002; @radeloff2019; @coops2019;
@razenkova2023]. Ecologically available energy is the amount of energy
produced in an ecosystem, which is sensitive to limitations such as
recent disturbances and the water and nutrient supplies, and is
measurable via productivity [@bonn2004; @nguyen2022]. This differs from
ambient energy in that it goes beyond the climatic conditions of an
area, and concerns the amount of resources available for consumption in
an ecosystem [@hawkins2003]. Clear links have been developed between
remote sensing-derived spectral reflectance, particularly in the
near-infrared and the red parts of the electromagnetic spectrum
[@rouse1974], which relate to pigment concentrations and leaf structure,
respectively, as well as productivity [@yoder1994]. Consequently, the
normalized difference vegetation index (NDVI) and other vegetation
indices have been correlated with productivity measures [@huang2021;
@sims2006]. The relationship however, is typically non-linear,
principally due to saturation in the spectral response at moderate to
high leaf area index values [@myneni1994]. As a result, the strength of
the relationship between spectral indices and productivity depends on
the ecosystem type, land cover type, amount of canopy cover, and
productivity in the region [@shi2017]. Likewise, relationships between
productivity indicators and biodiversity also contain uncertainty,
recognizing that biodiversity is governed by a variety of factors of
which productivity is but one [@skidmore2021]. Despite these
uncertainties, remotely sensed vegetation indices have enabled the
assessment of biodiversity patterns at single time points [@bonn2004;
@nagendra2010; @phillips2008], or through time when using time series
data [@berry2007; @radeloff2019; @coops2019]

@@

bec_tab %>%
  gt() %>%
  cols_label(
    group = "BEC Group",
    zone = "BEC Zone",
    zone_nm = "Full Name",
    ppt = "Average Monthly Precipitation (mm/yr)",
    tmax = "Average Monthly Max Temperature (°C)",
    tmin = "Average Monthly Min Temperature (°C)",
    n = "n"
  ) %>%
  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, monthly average climate values for precipitation, maximum temperature, and minimum temperature, and the number of samples taken from each BEC zone. Climate data from Wang et al. 2016 averaged for 1991-2020.")
```

@@

### Forest Structure

The forest structural attributes (canopy height, canopy cover,
structural complexity \[coefficient of variation of height returns\],
basal area, aboveground biomass, and gross stem volume) were retrieved
for 2015 from the 30-m spatial resolution dataset generated by Matasci
et al. [@matasci2018; @matasci2018b]. Their method extends structural
attributes derived from a set of lidar acquisitions and field plots
across Canada. These data were generated for the treed pixels of Canada
using the BAP Landsat surface reflectance composite [@hermosilla2016]
and auxiliary data (i.e., topography, geography) using a k-Nearest
Neighbour imputation [@matasci2018; @matasci2018b]. Accuracy metrics for
imputed forest structural attributes range from an RMSE of 29.7%
(structural complexity) to 82.3% (gross stem volume) and *R^2^* ranging
from 0.125 (structural complexity) to 0.712 (gross stem volume)
[@matasci2018; @matasci2018b].

The DHIs for the terrestrial area of BC were calculated following the
procedures described in Razenkova et al
@razenkovaMediumresolutionDynamicHabitatInPress. Although DHIs are
typically derived from high-temporal, coarse spatial resolution imagery
such as MODIS, we used Landsat surface reflectance data due to its finer
spatial resolution, which corresponds to the 30 m ALS derived forest
structure data. To account for the lower temporal resolution of Landsat,
we generated a synthetic year of monthly data from a ten-year timespan
of Landsat imagery. Google Earth Engine (GEE) [@gorelick2017] was used
to obtain all valid Landsat pixels for the study area over a ten-year
period centered on 2015. Selected pixels were filtered for shadows,
clouds, and cloud shadows using the QA band derived from the fmask
algorithm [@zhu2012], available on GEE. The NDVI was then calculated for
each pixel in each image and summarized into a synthetic year of monthly
data by calculating the median of each month's NDVI values, disregarding
the year of the image. The calculations for the Cumulative DHI, Minimum
DHI, and Variation DHI can be found in Table 2, alongside a summary of
the main datasets analyzed in this study.

### Ancillary Data

@@

The forest structure, forest disturbance, and land cover layers are
based on a 30-m best-available-pixel (BAP) composite generated by
Hermosilla et al. @hermosilla2016 using the Composite2Change approach.
The composites were developed for each year from 1984 to 2019 by
selecting the best available imagery from the available growing season
Landsat observations, removing pixels with clouds, cloud shadows, and
haze. The methodology uses the scoring method from White et al.
@white2014 to select the best pixel for a given year from Landsat-5
Thematic Mapper, Landsat-7 Enhanced Thematic Mapper Plus, and Landsat-8
Operational Land Imager imagery. A spectral trend analysis was conducted
over these initial BAP composites using the Normalized Burn Ratio (a
vegetation index) on each pixel, to remove unscreened noise, detect
changes, and temporally interpolate data gaps. This resulted in a
gap-free surface-reflectance composite across Canada [@hermosilla2015].
The forest structure and land cover datasets are based on BAP data from
2015, while the forest disturbance mask covers the period from
1985-2020.

@@

@@

To obtain representative sample units of the EBV metrics across the
study area, we implemented stratified random sampling across the union
of BEC zones (Table 1) and forest types (i.e., coniferous, broadleaf,
mixed wood, and wetland treed; see @sec-study-area) of BC. Sample units
were restricted to forested pixels that were surrounded by the same
forest type to reduce the chance of sampling mixed pixels. Sample units
were selected from homogeneous areas to reduce uncertainty associated
with the input datasets [@shang2020]. Homogeneous areas were defined as
3x3-pixel windows with a coefficient of variation for canopy cover and
canopy height below 0.5. A minimum sampling distance of 1 km was
implemented to reduce the effects of spatial autocorrelation. Sample
units that were disturbed in the last 35 years were excluded by using
the disturbance mask generated for the forested ecosystems of Canada
[@hermosilla2016]. A maximum of 500 sample units were selected from each
stratum. In the strata where this was not achieved, all sampling units
meeting the above restrictions were selected (see Table 1). The sampling
was conducted in R version 4.2.2 [@R-base] using the **sgsR** package
[@R-sgsR]. Neighbourhood analyses for the land cover classes and
coefficient of variations of canopy height and cover were calculated in
Python version 3.9.

@@

Variation partitioning is displayed using a Venn diagram in which the
percentage of variance explained by each dataset is in a circle, and the
overlap between circles represents the overlap in variance explained. We
add paths to the Venn diagram to indicate how the structural attributes
contribute the variation in the DHIs. Because the variation partitioning
results summarize the full multivariate solution, rather than individual
RDA axes, we present variation partitioning results alongside the path
diagrams illustrating the details of the RDA results. RDA and variation
partitioning calculations were done in R [@R-base] version 4.2.2 using
the **vegan** package [@R-vegan]. The code associated with the
processing and analysis is available at
https://github.com/emuise/code-structProdSem.

We present a simplified flow diagram of the data provenance to result
pipeline in @fig-flow.

```{r}
#| label: fig-flow
#| fig-cap: Flow diagram of the data to indicators to results pipeline implemented in this study.

knitr::include_graphics(here::here("outputs", "p2_flow.drawio.png"))
```

@@

Overall in BC, the intra-annual patterns of productivity as represented
by the DHI variables were weakly related to forest structure
(@fig-rda-var). Less than 16% of the overall variation in the DHIs was
explained by the forest structural attributes. Due to strong
correlations between the axes of the DHIs, the majority of the variation
was held within the first RDA axis, which had strong loadings for all
DHIs, with positive loadings for the cumulative and minimum DHIs, and
negative loadings for the variation DHI. As such, we identified the RDA1
axis as representing overall productivity. The overall productivity axis
had similar loadings (between 0.26-0.28) for all modelled forest
structural attributes and canopy cover. Canopy height had a smaller
loading (0.13), while structural complexity was negatively correlated
with the overall productivity axis. RDA2 explained much less of the
variation in the DHIs. RDA2 had a large negative loading on the minimum
DHI (-0.47), with smaller, positive loadings on the cumulative (0.26)
and variation (0.29) DHIs. Due to the minimum DHI having the largest
influence on RDA2, we identified it as a seasonality axis. Canopy cover
and structural complexity had the strongest (0.18 and 0.08,
respectively) loadings on the seasonality axis (@fig-rda-var A).
Variation partitioning highlighted that the majority of the explained
variation in the DHIs was due to the primary forest structural
attributes (canopy cover, canopy height, and structural complexity; 7.7%
variation explained). The overlap between primary and modelled
attributes was 4.2%, and the modelled attributes explained 3.4% of the
variation on their own (@fig-rda-var B).

@@

By forest type, results indicated similar or increased amounts of
variance explained when compared to the overall dataset
(@fig-fracts-bar). By BEC zone, results generally indicated smaller
amounts of variation explained, which may be due to the relatively small
environmental gradients within the zones. The overlap between the
primary and modelled datasets was often responsible for the majority of
the variation explained in the DHIs, except when stratifying by forest
types, where this was only true for coniferous forests. For broadleaf,
mixed wood, and wetland-treed forest types most of the explained
variation in the DHIs was contributed by the primary forest structural
attributes. A similar pattern was found by BEC zones, where the
variation in the DHIs was principally explained by the overlap between
primary and modelled forest structural attributes, except for CWH, where
the variation was entirely explained by the primary forest structural
attributes. The Coastal BEC group had the lowest amount of variation
explained, with Coastal Douglas-fir having \<1% of variation explained
by the structural data. Overall, the DHIs were decoupled from the forest
structural attributes, as shown by the overall variation explained being
under 32% regardless of data stratification.

@@

Stratification by forest type and ecosystem indicated varying loadings
between the RDA axes and predictor/response variables (@fig-radar). All
DHI variables had high loadings on RDA1 (similar to the overall
dataset), with positive cumulative and minimum DHI loadings, and
negative variation DHI loadings. Colder BEC groups (Montane, North, and
Alpine) had smaller minimum DHI loadings, which were half as large as
the other DHI loadings. As such, we identify the first axis as being
consistently associated with overall productivity. The predictor
loadings for canopy cover and the modelled forest structural attributes
were generally similar within the same stratification/BEC groups. The
mixed wood forest and Coastal BEC group had no significant loadings from
modelled forest structural attributes for RDA1. Examining the predictor
loadings for RDA1 spatially across the province shows a geographic
pattern in the loading strengths of primary forest structural attributes
(@fig-fcc A). The interior (Northern and Southern BEC groups) of the
province generally showed high canopy cover loadings, while the Boreal
zones (Boreal White and Black Spruce and Sub-Boreal Pine – Spruce) in
the northwest had large canopy height loadings. Coastal zones were
driven similarly by canopy height and structural complexity, with very
low or non-significant loadings from canopy cover (@fig-fcc A;
@fig-radar).

@@

Only four of sixteen BEC zones had a second RDA axis, with none being
found in Coastal or Alpine BEC zones (@fig-radar; @fig-fcc). Two Montane
BEC zones had a secondary axis, while the Northern and Southern interior
groups each had one zone with a secondary axis. RDA2 often showed the
largest absolute loadings in the minimum DHI, which was frequently, but
not always, negative (@fig-radar; see supplementary material). The
Montane group had the largest absolute minimum and variation DHI
loadings, which were notably the opposite sign of all other groups with
a secondary axis. RDA2 had lower absolute predictor loadings (maximum of
\~0.25) when compared to RDA1 (maximum of \~0.5), with infrequently
significant loadings for canopy cover and structural complexity
(@fig-radar). There was no consistent loading pattern found in RDA2 for
the primary predictor variables (@fig-radar; @fig-fcc C). Similar to
RDA1, the modelled forest structural attribute loadings were generally
similar to one another (@fig-radar; @fig-fcc D).

@@

Relationships between forest ecosystem structure and function are being
examined more frequently due to advances in data collection methods,
such as remote sensing [@ali2019; @radeloff2019; @ali2016]. This paper
assesses the complementarity of two potential ecosystem-scale EBV
indicators for ecosystem monitoring across BC, Canada. Redundancy
analysis was used to determine the information overlap of wall-to-wall
maps of forest structural attributes and the DHIs, which were both
generated from Landsat imagery. This analysis revealed a weak
relationship between the two sets of remote sensing derived indicators,
with only 15.7% of the variation in the DHIs being explained by the
forest structural attributes (@fig-rda-var). This relationship was also
assessed across the different forest classes and ecosystems of British
Columbia, and similar results were found (@fig-fracts-bar). The small
amount of overlap between the two datasets suggests that they are
suitable for use in ecosystem monitoring programs, despite both being
derived from Landsat imagery.

We also evaluated whether primary forest structural attributes provide
more information than modelled forest structural attributes for
selecting indicators within a single EBV class. It was found that
standard lidar-derived measurements, such as canopy height, canopy
cover, and structural complexity (as proposed by Valbuena et al.
[@valbuena2020], contribute the majority of the explained variation in
the DHIs, either independently or in their overlap with the modelled
forest structural attributes. This could be expected as modelled forest
structural attributes are generally directly based on the primary forest
structural attributes [@coops2021], however, the creation of modelled
forest structural attributes involves modelling expertise and processing
time overhead [@matasci2018; @matasci2018b; @duncanson2022]. One
advantage of interpreted products is their potential uptake by a broader
set of users and their ability to be closely linked with management
actions and monitoring results. This is exemplified by the use of forest
biomass monitoring for climate change mitigation [@duncanson2022]. While
either set of forest structural attributes may be relevant for a given
monitoring task, it is important to note that the primary structural
attributes contain the majority of the variation across both classes of
structural attributes (@fig-fracts-bar). Additionally, the modelled
structural attributes are generally not related to structural
complexity, which has often been found to be strongly related to
biodiversity metrics [@ehbrecht2021]. Therefore, we generally recommend
using the primary forest structural attributes.

Our findings indicate that remote sensing-derived forest ecosystem
structure and forest ecosystem function EBVs can be applied in a
complementary fashion for ecosystem monitoring [@hansen2021], despite
both EBVs using surface reflectance values from the Landsat series of
satellites in either their calculation or imputation [@radeloff2019;
@matasci2018; @matasci2018b]. While we assessed the complementarity of
ecosystem structure and function, a single pair of EBVs measurable at
the ecosystem scale, other EBV class combinations at finer scales should
also be assessed for complementarity. This is especially relevant when
remote sensing datasets such as Landsat can generate a large number of
EBVs in different classes [@skidmore2021]. In addition, Skidmore et al.
[@skidmore2021] highlighted that many satellite-derived biodiversity
products can inform upon multiple EBV classes simultaneously -- e.g.,
leaf area index being recognized as a ecosystem structure, ecosystem
function, and species trait EBV indicator -- depending on the spatial
scale of analysis and goal of the monitoring program. As such, it is
relevant to consider which EBV classes biodiversity products can inform,
for given spatial and temporal extents. Other EBVs, such as those found
in the community composition and species population classes may also be
informed by these products. Further, their reliance on species level
data also means that they could be inherently correlated with one
another.

@@

Forest types with deciduous trees (namely broadleaf and mixed wood) had
large, negative structural complexity loadings when compared to the
other forest structural attributes (@fig-radar), with forest types
having no modelled structural attributes associated with the DHIs. This
structural complexity-DHIs linkage may be due to understory cover
providing additional photosynthesis in these forest types
[@nilsson2005]. In coniferous forests, on the other hand, the DHIs show
a weaker relationship with structural complexity than all other forest
structural attributes. Notably, the Alpine, Montane, and Northern BEC
groups show decoupling of the three DHIs (@fig-radar), with lower
loadings between the Minimum DHI and RDA1. This decoupling in the DHIs
is potentially due to the near certainty of a pixel containing snow
cover over the course of a year leading to consistently low minimum DHI
values across the ecosystem, and thus to weaker minimum DHI loadings
[@razenkova2020; @beck2006]. These ecosystems with DHI decoupling also
showed relatively strong relationships to canopy cover and weak
relationships with structural complexity. This weak relationship to
structural complexity may be due to the dominance of coniferous forests
in these ecosystems, which have been shown to have lower forest
structural complexity across the globe [@ehbrecht2021], with weaker
relationships between canopy height and structural complexity being
found than deciduous forests [@atkins2022].
